Attribution of undesirable character traits, rather than trait-based dehumanization, predicts punishment decisions

### 2024-07-03 manuscript.Rmd
Update manuscript.Rmd

Understanding the motivations that lead to intergroup harm has been a driving force behind social psychology since its conception [@Farr1996; @Gaines1995]. Many psychological processes have been shown to exacerbate outgroup derogation, including prejudicial attitudes and stereotyping [@Dovidio2010; @Major2009]. Over the last 25 years, an increasing body of research has investigated the extent to which a psychological process of dehumanization increases the risk of intergroup harm [@Bloom2017; @Bloom2022; @Giner-Sorolla2021; @Goldenberg2021; @Lang2010; @Lang2020; @Manne2016; @Manne2018; @Over2021a; @Over2021b; @Rai2017; @Ruiter2022; @Smith2011; @Smith2020; @Smith2021; @Vaes2021]. According to the dehumanization hypothesis, when members of an outgroup are perceived as less human than ingroup members, they are at greater risk of harm [@Haslam2019; @Haslam2014; @Haslam2015; @Smith2011; @Smith2016]. Subtle forms of dehumanization are thought to be pervasive in contemporary society. For example, the dehumanization of national groups [@Bain2011; @Leyens2003], religious groups [@Banton2020; @Leidner2013], individuals on a low income [@Loughnan2013], and refugees [@Bruneau2018; @Esses2008] has been reported in the literature.

@@

@Bastian2013 conducted influential empirical studies testing the hypothesised association between the denial of human character traits and the endorsement of harsh punishment. The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).

>>>>

Understanding the motivations that lead to intergroup harm has been a driving force behind social psychology since its conception [@Farr1996; @Gaines1995]. Many psychological processes have been shown to exacerbate outgroup derogation, including prejudicial attitudes and stereotyping [@Dovidio2010; @Major2009]. Over the last 25 years, an increasing body of research has investigated the extent to which a psychological process of dehumanization increases the risk of intergroup harm [@Bloom2017; @Bloom2022; @Giner-Sorolla2021; @Goldenberg2021; @Lang2010; @Lang2020; @Manne2016; @Manne2018; @Over2021a; @Over2021b; @Rai2017; @Ruiter2022; @Smith2011; @Smith2020; @Smith2021; @Vaes2021]. According to the dehumanization hypothesis, when members of an outgroup are perceived as less human than ingroup members, they are at greater risk of harm [@Haslam2019; @Haslam2014; @Haslam2016; @Over2021a; @Smith2011; @Smith2016]. Subtle forms of dehumanization are thought to be pervasive in contemporary society. For example, the dehumanization of national groups [@Bain2011; @Leyens2003], religious groups [@Banton2020; @Leidner2013], individuals on a low income [@Loughnan2013], and refugees [@Bruneau2018; @Esses2008] has been reported in the literature.

@@

@Bastian2013 conducted influential empirical studies testing the hypothesised association between the denial of human character traits and the endorsement of harsh punishment (see also @Barber2022; @Chen-Xia2023; Kasper2022; Morehouse2023; Rousseau;2023; West2023). The researchers measured how trait-based dehumanization influenced participants' punishment of criminals. Participants were asked to rate their agreement with four items assessing animalistic dehumanization of criminals: "*I felt like the person in the story was refined and cultured*" [reversed], "*I felt like the person in the story was rational and logical, like they were intelligent*" [reversed] "*I felt like the person in the story lacked self-restraint, like an animal*", and "*I felt like the person in the story was unsophisticated*". Participants were also asked to rate their agreement with four items assessing mechanistic dehumanization of criminals: "*I felt like the person in the story was open minded, like they could think clearly about things*" [reversed], "*I felt like the person in the story was emotional, like they were responsive and warm*" [reversed], "*I felt like the person in the story was superficial like they had no depth*", "*I felt like the person in the story was mechanical and cold, like a robot*". Bastian and colleagues reported that both forms of dehumanization predicted endorsement of harsh punishment for the criminals portrayed in their stimuli, concluding that their participants viewed criminals as '*subhuman and beastly*' (pp.9).
